Supplemental Material for ``Assessing United States county-level exposure for research on tropical cyclones and human health"

### 2019-07-28 writing/ehp_discussion.tex
Work more on paper draft

The flood and tornado data came from the \ac{NOAA} Storm Events database,
which, while a widely-used database of events maintained by \ac{NOAA}, is based
on reports, and so may be prone to underreporting~\parencite{Ashley2008flood,
Curran2000}, especially in less populated areas~\parencite{Witt1998,
Ashley2007}, as well as to other reporting errors. While these are important
limitations, we selected these data sources as among the best currently
available for measuring each of these hazards at a multi-county, multi-year
scale.

@@

Dissimilarity in exposure to specific hazards of tropical cyclones can result
in differing patterns in estimates of county-level exposure
(Figure~\ref{fig:ivanexposure}), which can lead to substantial differences
in estimates of population physical exposure
(Figure~\ref{fig:topelecdependexposure}), based on which storm hazard is
considered. The measure of a hurricane's intensity, the Saffir-Simpson scale,
expresses intensity as a function of wind speed, and so fails to capture
potential risks from other hazards of the storm~\parencite{smith2009}, and
previous assessments of physical exposure to tropical cyclones have often
focused on winds or distance from the storm. For example, a global study of
exposure and vulnerability to natural hazards used a combination of windspeed
above a certain threshold and distance from the storm's track to assess
exposure~\parencite{peduzzi2009assessing}. While many storm impacts might be
most strongly linked to wind hazards, there is a growing recognition of the
potential risks of adverse health outcomes and property damage from rain- and
flood-related tropical cyclone hazards in the \ac{US}~\parencite{smith2009}.
Such assessments may underemphasize potentially dangerous exposures to certain
regions of the \ac{US}, particularly inland locations. In fact, based on our
assessment of physical exposures to tropical cyclones among
electricity-dependent Medicare beneficiaries, the highest expected physical
exposure under any metric was measured for Philadelphia County, PA, under the
flood metric (Figure~\ref{fig:topelecdependexposure}).  While this county has a
similar-sized population of electricity-dependent Medicare beneficiaries as
Miami-Dade County, FL (Figure~\ref{fig:topelecdependexposure}), the frequency
of storm-related flood events in this inland county results in a very high
expected physical exposure among this subpopulation, while this county is not
included in the top ten counties for physical exposure based on most of the
other exposure metrics.

>>>>

For the flooding events data, we found it often corresponded with measurements
from \ac{US} Geographical Survey streamgages, with flood event listings more
likely during storms which elevated stream flows. However, there are
differences between the two flooding datasets, and these highlight both the
difficulty of measuring flood exposure at the county level and inherent
challenges in using the storm event database for epidemiologic exposure
assessment.  The flood and tornado data came from the \ac{NOAA} Storm Events
database, which, while a widely-used database of events maintained by
\ac{NOAA}, is based on reports, and so may be prone to
underreporting~\parencite{Ashley2008flood, Curran2000}, especially in less
populated areas~\parencite{Witt1998, Ashley2007}, as well as to other reporting
errors.

\begin{comment}
[For discussion?] Based on these results, the two data sources were generally
in agreement, but the analysis highlights the difficulty of assessing
storm-related flooding at a multi-county, multi-year scale, as flooding can be
very localized within a county or can \ldots .   
\end{comment}

While these are important caveats for the data, we selected these data sources
as among the best currently available for measuring each of these hazards at a
multi-county, multi-year scale.

@@

The conducted case study demonstrates how differences in exposure assessment
across storm exposure metrics can result in differing estimates of expected
physical exposure to tropical cyclones among a population
(Figure~\ref{fig:topelecdependexposure}).  Previous assessments of physical
exposure to tropical cyclones have often focused on solely storm winds and
storm proximity. For example, a global study of exposure and vulnerability to
natural hazards used a combination of windspeed above a certain threshold and
distance from the storm's track to assess
exposure~\parencite{peduzzi2009assessing}. While many storm impacts might be
most strongly linked to wind hazards, there is a growing recognition of the
potential risks of adverse health outcomes and property damage from rain- and
flood-related tropical cyclone hazards in the \ac{US}~\parencite{smith2009}.
Such assessments may underemphasize potentially dangerous exposures to certain
regions of the \ac{US}, particularly inland locations. In fact, based on our
assessment of physical exposures to tropical cyclones among
electricity-dependent Medicare beneficiaries, the highest expected physical
exposure under any metric was measured for Philadelphia County, PA, under the
flood metric (Figure~\ref{fig:topelecdependexposure}).  While this county has a
similar-sized population of electricity-dependent Medicare beneficiaries as
Miami-Dade County, FL (Figure~\ref{fig:topelecdependexposure}), the frequency
of storm-related flood events in this inland county results in a very high
expected physical exposure among this subpopulation, while this county is not
included in the top ten counties for physical exposure based on most of the
other exposure metrics.

### 2019-07-28 writing/ehp_figures.tex
Work more on paper draft

\begin{figure}[tbhp!]
\centering
\includegraphics[width=0.8\linewidth]{floodcomparison}
\caption{Comparison for a sample of study counties of NOAA Storm Events
listings versus five-day total streamflows at county streamflow gages during
tropical storms. The number of streamflow gages used in analysis are given in
parentheses beside the county's name in the panel title. Each point represents
a single tropical storm. Note that the x-axis scales differ by county,
depending on the number of streamflow gages and typical flow rates for each
gage, and are on a log-10 scale.}
\label{fig:floodcomparison}
\end{figure}

@@

\begin{figure}%[tbhp] 
\centering
\includegraphics[width=15cm]{averageexposureonly.pdf} 
\caption{Average number of storm exposures per decade in U.S. counties for each
exposure metric. The criteria behind each of the five metrics is given in Table
\ref{tab:exposuremetrics}. The years used to estimate these averages are based
on years of available exposure data (distance and wind:~1988\,--\,2015; rain:
1988\,--\,2011; flood and tornado:~1996\,--\,2015). } 
\label{fig:averageexposure} 
\end{figure}

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=15.5cm]{topelecdependexposure}
\caption{Counties with the highest estimated physical exposure among
electricity-dependent Medicare beneficiaries exposed to storms per year in U.S.
counties for each exposure metric. The color of each bar indicates the number
of Medicare beneficiaries in the county reliant on electricity-dependent
medical and assistive equipment as of July~2017. The length
of each bar shows the average expected number of these electricity-dependent
Medicare beneficiaries exposed to tropical storms per year based on a given
exposure metric.}
\label{fig:topelecdependexposure}
\end{figure*}

>>>>

\begin{figure}[tbhp!]
\centering
\includegraphics[width=0.8\linewidth]{floodcomparison}
\caption{Comparison for a sample of study counties of NOAA Storm Events 
	listings versus five-day total streamflows at county streamflow 
	gages during tropical cyclones. Each small plot shows results for 
	one of the sample counties. Each point represents a single tropical 
	storm, and the point's position along the x-axis shows the highest 
	daily total streamflow (cubic feet per second), summed across all 
	identified streamgages in the county, for a five-day window centered 
	on the day of the storm's closest approach to the county. The y-axis 
	separates storms for which a flood event was reported in NOAA's Storm Events 
	database for the county with a start date within the five-day window 
	of the storm's closest approach to the county. The color of each point 
	gives the percent of streamflow gages in the county with a daily streamflow 
	that exceeded a threshold of flooding on any day during the five-day window.
	Comparison for a sample of study counties of NOAA Storm Events 
	listings versus five-day total streamflows at county streamflow gages during 
	tropical storms. The number of streamflow gages used in analysis are given in 
	parentheses beside the county's name in the panel title. Each point represents 
	a single tropical storm. Note that the x-axis scales differ by county, 
	depending on the number of streamflow gages and typical flow rates for each 
	gage, and are on a log-10 scale.}
\label{fig:floodcomparison}
\end{figure}

@@

\begin{figure}%[tbhp] 
\centering
\includegraphics[width=15cm]{averageexposureonly.pdf} 
\caption{Average number of storm exposures per decade in U.S. counties for each 
	exposure metric. The criteria behind each of the five metrics is given in 
	Table \ref{tab:exposuremetrics}. The years used to estimate these averages 
	are based on years of available exposure data 
	(distance and wind:~1988\,--\,2015; rain: 1988\,--\,2011; 
	flood and tornado events:~1996\,--\,2015). } 
\label{fig:averageexposure} 
\end{figure}

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=15.5cm]{topelecdependexposure}
\caption{Study counties with the highest expected physical exposure per year among
	 electricity-dependent Medicare beneficiaries for each exposure metric. 
	 The color of each bar indicates the number of Medicare beneficiaries in the 
	 county reliant on electricity-dependent medical and assistive equipment as 
	 of July~2017. The length of each bar shows the average expected physical exposure
	 to tropical cyclones among this population based on a given exposure metric, i.e., 
	 the expected number of these electricity-dependent Medicare beneficiaries exposed 
	 to tropical storms per year based on that metric.}
\label{fig:topelecdependexposure}
\end{figure*}

### 2019-07-28 writing/ehp_methods.tex
Work more on paper draft

To identify flood- and tornado-based exposures to tropical cyclones in \ac{US}
counties, we used event listings from the National Oceanic and Atmospheric
Administration (NOAA)'s Storm Event Database~\parencite{stormevents}. For each
tropical cyclone, we identified all events with event types related to flooding
(``Flood", ``Flash Flood", ``Coastal Flood") and tornadoes (``Tornado") and
that occurred in a county within~500 \si{\kilo\metre} of the tropical cyclone's
track and with a start date within a five-day window centered on the date of
the tropical cyclone's closest approach to the
county~\parencite{hurricaneexposuredata}. ``Flood", ``Flash Flood" and
``Tornado" events in this database were reported by county \ac{FIPS} code.
``Coastal Flood" events were reported by forecast zone; for these, the event
was matched to the appropriate county if possible using regular expression
matching of listed county names~\parencite{noaastormevents}. While this
database has recorded storm data since~1950, its coverage changed substantially
in~1996 to cover a wider variety of storm events~\parencite{stormevents}. We
therefore only considered these metrics of hurricane exposure for tropical
cyclones in~1996 and later.

The tornado observations from this dataset form the traditional tornado
tracking dataset for the \ac{US} [ref for this?], and so we did not conduct
analyses to validate the tornado event data.  As validation for the
flood data, we investigated a sample of study counties, comparing the flood
status, based on NOAA Storm Events listings~\parencite{stormevents,
noaastormevents}, with streamflow measurements at \ac{US} Geological Survey county
streamflow gages during tropical storms~\parencite{usgsgages, countyfloods,
dataRetrieval}.  We considered nine study counties, selecting counties
geographically spread through storm-prone areas of the eastern \ac{US} and with
multiple streamflow gages reporting data during events (Baltimore County, MD;
Bergen County, NJ; Escambia County, FL; Fairfield County, CT; Fulton County,
GA; Harris County, TX; Mobile County, AL; Montgomery County, PL; and Wake
County, NC). For each county, we first identified all streamflow gages in the
county that had complete data for Jan.~1~1996\,--\,Dec.~31,~2015. For each
tropical cyclone that came within~500~\si{\kilo\metre} of one of these counties
between~1996 and~2015, we calculated the highest daily total streamflow (cubic
feet / second), summed across all identified streamgages in the county, for a
five-day window centered on the day of the storm's closest approach to the
county.  We also calculated the percent of streamflow gages in the county with
a daily streamflow that exceeded a threshold of flooding (the streamgage's
median value for annual peak flow~\parencite{countyfloods}) on any day during
the five-day window. We compared these measurements to each county's flood
status based on NOAA Storm Events listings for each tropical cyclone that came
within~500~\si{\kilo\metre} between~1996 and~2015, to explore if flood event
listings in the NOAA Storm Events database concurrent with tropical cyclones
were associated with higher streamflows at gages within the county.

>>>>

To identify flood- and tornado-based tropical cyclone exposures in \ac{US}
counties, we used event listings from the National Oceanic and Atmospheric
Administration (NOAA)'s Storm Event Database~\parencite{stormevents}. For each
tropical cyclone, we identified all events with event types related to flooding
(``Flood", ``Flash Flood", ``Coastal Flood") and tornadoes (``Tornado") and
that occurred in a county within~500 \si{\kilo\metre} of the tropical cyclone's
track and with a start date within a five-day window centered on the date of
the tropical cyclone's closest approach to the
county~\parencite{hurricaneexposuredata}. ``Flood", ``Flash Flood" and
``Tornado" events in this database were reported by county \ac{FIPS} code and
so could be directly linked to counties.  ``Coastal Flood" events were reported
by forecast zone; for these, the event was matched to the appropriate county if
possible using regular expression matching of listed county
names~\parencite{noaastormevents}. While this database has recorded storm data
since~1950, its coverage changed substantially in~1996 to cover a wider variety
of storm events~\parencite{stormevents}. We therefore only considered these
metrics of hurricane exposure for tropical cyclones in~1996 and later.

The tornado observations from this dataset form the traditional tornado event
database for the \ac{US}, and so we did not conduct analyses to validate the
tornado event data. It is difficult to characterize flooding at the county
level because flooding can be very localized and can be triggered by a variety
of causes. To investigate the extent to which the NOAA flood event data
captures extremes that might be identified with other flooding data sources, we
investigated a sample of study counties, comparing the flood event data during
tropical storms with streamflow measurements at \ac{US} Geological Survey
county streamflow gages \parencite{usgsgages, countyfloods, dataRetrieval}.

We considered nine study counties, selecting counties geographically spread
through storm-prone areas of the eastern \ac{US} and with multiple streamflow
gages reporting data during events (Baltimore County, MD; Bergen County, NJ;
Escambia County, FL; Fairfield County, CT; Fulton County, GA; Harris County,
TX; Mobile County, AL; Montgomery County, PL; and Wake County, NC). For each
county, we first identified all streamflow gages in the county that had
complete data for Jan.~1~1996\,--\,Dec.~31,~2015. If a storm did not come
within~500~\si{\kilo\metre} of a county, it was excluded from this analysis,
but all other study storms were considered. For each storm and county, we
summed daily total streamflow across all streamgages for each day in the
five-day window around the storm's closest approach. We took the maximum daily
streamflow total during those five days as a measure of the county's maximum
streamflow during that storm. We also calculated the percent of streamflow
gages in the county with a daily streamflow that exceeded a threshold of
flooding (the streamgage's median value for annual peak
flow~\parencite{countyfloods}) on any day during the five-day window. We
investigate how these measurements varied between storms with associated flood
events in the NOAA Storm Events data versus storms without an event listing, to
explore if storms with flood event listings tended to be  associated with
higher streamflows at gages within the county.

### 2019-07-28 writing/ehp_results.tex
Work more on paper draft

\paragraph{Flood data validation}

For flood data, we compared the flood status measurements, based on event
listings in the NOAA Storm Events database, to streamflow measurements at
\ac{US} Geological Survey gages within nine sample counties
(Figure~\ref{fig:floodcomparison}). Each small plot in
Figure~\ref{fig:floodcomparison} shows results for one of the sample counties.
Each point represents a single tropical storm, and the point's position along
the x-axis shows the highest daily total streamflow (cubic feet per second),
summed across all identified streamgages in the county, for a five-day window
centered on the day of the storm's closest approach to the county. The y-axis
separates storms for which a flood event was reported in NOAA's Storm Events
database for the county with a start date within the five-day window of the
storm's closest approach to the county. The color of each point gives the
percent of streamflow gages in the county with a daily streamflow that exceeded
a threshold of flooding on any day during the five-day window.  Stream gage
data generally indicated higher discharge in these counties on dates identified
as flood events, particularly in counties with more reporting stream gages
(e.g., Harris County, TX, with~10 reporting stream gages;
Figure~\ref{fig:floodcomparison}). There were some cases, however, where the
two flooding data sources were not completely consistent. 
For example, there
were one or two tropical cyclones in several of the counties (Mobile County,
AL, Escambia County, FL, and Montgomery County, PA) which were classified as
flood events based on NOAA storm event listings but for which the total
discharge across county streamflow gages was below the 25\textsuperscript{th}
percentile of measurements during storms without a flood event listing. 
For tropical cyclones that did not have an associated flood listing in 
the NOAA Storm Events database, most had lower total streamflow discharge
across county gages compareed to storms with flood event listings, and in all
cases but one none of the gages were over the flooding threshold. The exception
was \ldots in Fulton County, GA. [More on which storm and why---I think there
was already a flood event that had started earlier, so it missed the cutoff for
the startdate being within the range.]

[For discussion?] Based on these results, the two data sources were generally
in agreement, but the analysis highlights the difficulty of assessing
storm-related flooding at a multi-county, multi-year scale, as flooding can be
very localized within a county or can \ldots .

@@

Finally, we conducted a case study to investigate how these differences in
exposure assessment across the exposure metrics could affect estimates of
expected physical exposure among a vulnerable subpopulation,
electricity-dependent Medicare beneficiaries.
Figure~\ref{fig:topelecdependexposure} shows the study counties with the
highest expected physical exposures in this subpopulation based on each
exposure metric. A few counties (Miami-Dade County, FL; Harris County, TX) were
ranked in the top three counties of physical exposure for almost all exposure
metrics (Figure~\ref{fig:topelecdependexposure}). However, there were key
differences when using different exposure metrics. For example, Florida
counties made up most of the counties counties with highest expected physical
exposure based on wind- and tornado-based metrics, while over half of the
counties with highest expected physical exposures for the flood-based metric
were in Mid-Atlantic states (PA, NY).

>>>>

\paragraph{Flood data analysis}

For flood data, we compared flood status, based on event listings in the NOAA
Storm Events database, to streamflow measurements at \ac{US} Geological Survey
gages within nine sample counties (Figure~\ref{fig:floodcomparison}). Each
small plot in Figure~\ref{fig:floodcomparison} shows results for one of the
sample counties.  Each point represents a single tropical storm, and the
point's position along the x-axis shows the highest daily total streamflow
(cubic feet per second), summed across all identified streamgages in the
county, for a five-day window centered on the day of the storm's closest
approach to the county. The y-axis separates storms for which a flood event was
reported in NOAA's Storm Events database for the county with a start date
within the five-day window of the storm's closest approach to the county. The
color of each point gives the percent of streamflow gages in the county with a
daily streamflow that exceeded a threshold of flooding on any day during the
five-day window.

For flood data, we compared flood status, based on event listings in the NOAA
Storm Events database, to streamflow measurements at \ac{US} Geological Survey
gages within nine sample counties (Figure~\ref{fig:floodcomparison}).  Across
the sampled counties, streamgage data generally indicated higher discharge in
on dates identified as flood events, particularly in counties with more
reporting stream gages (e.g., Harris County, TX, with~10 reporting stream
gages; Figure~\ref{fig:floodcomparison}). There were some cases, however, where
the two flooding data sources were somewhat inconsistent.  For example, there
were one or two tropical cyclones in several of the counties (Mobile County,
AL, Escambia County, FL, and Montgomery County, PA) which were classified as
flood events based on NOAA storm event listings but for which the total
discharge across county streamflow gages was below the 25\textsuperscript{th}
percentile of measurements taken during storms that did not have a flood event
listing. For tropical cyclones that did not have an associated flood listing in
the NOAA Storm Events database, most had lower total streamflow discharge
across county gages compared to storms with flood event listings, and in all
cases but one all gages were below the flooding threshold.  The exception was
one storm in Fulton County, GA (Ida in 2009). [More on which storm and why---I
think there was already a flood event that had started earlier, so it missed
the cutoff for the startdate being within the range.]

@@

Finally, we conducted a case study to investigate how these differences in
exposure assessment across the exposure metrics could affect estimates of
expected physical exposure among a vulnerable subpopulation,
electricity-dependent Medicare beneficiaries.
Figure~\ref{fig:topelecdependexposure} shows the study counties with the
highest expected physical exposures in this subpopulation based on each
exposure metric. A few counties (Miami-Dade County, FL; Harris County, TX) were
ranked in the top three counties for expected physical exposure in this
subpopulation for almost all exposure metrics
(Figure~\ref{fig:topelecdependexposure}). However, there were key differences
across exposure metrics. For example, most of the top counties based on wind-
and tornado-based metrics were in Florida, while over half of the counties with
highest expected physical exposures for the flood-based metric were in
Mid-Atlantic states (PA, NY).
